\Ace: Growing a Statically-Typed Language Inside a Python

### 2013-11-20 ace-pldi14/ace-pldi14.tex
Revisions of PLDI14 for speaking skills

%\category{D.3.2}{Programming Languages}{Language Classifications}[Extensible Languages]
%\category{D.3.4}{Programming Languages}{Processors}[Compilers]
%\category{F.3.1}{Logics \& Meanings of Programs}{Specifying and Verifying and Reasoning about Programs}[Specification Techniques]
%\keywords
%type-level computation, typed compilation
\section{Introduction}
Asking programmers to import a new library is far more practical than asking them to  adopt a new programming language. Indeed, recent empirical studies underscore  the difficulties of driving new languages into adoption, finding that extrinsic factors like compatibility with existing codebases and libraries, team familiarity and tool support are at least as important as intrinsic factors \cite{Meyerovich:2013:EAP:2509136.2509515,chen05,nguyen2010survey}. As a result, many developers cannot use  abstractions they might prefer because these abstractions are only available in languages they cannot adopt \cite{Meyerovich:2012:SDR:2414721.2414724,Meyerovich:2013:EAP:2509136.2509515}. This issue was perhaps most succinctly expressed by a participant in a recent study by Basili et al. \cite{basili2008understanding} who stated ``I hate MPI, I hate C++. [But] if I had to choose again, I would probably choose the same.''

@@

\subsection{Recursive Labeled Sums}
\begin{codelisting}
\lstinputlisting{datatypes_t.py}
\caption{\texttt{[datatypes\_t.py]} An example using statically-typed functional datatypes.}
\label{datatypest}
\end{codelisting}
\begin{codelisting}
\lstinputlisting{datatypes.py}
\caption{\texttt{[datatypes.py]} The dynamically-typed Python code generated by running \texttt{acec datatypes\_t.py}.}
\label{datatypes}
\end{codelisting}
Listing \ref{datatypest} shows an example of the use of statically-typed functional datatypes (a.k.a. recursive labeled sum types) \todo{change code to take a lambda} together with the Python implementation just described. It shows two syntactic conveniences that were not mentioned previously: 1) if no target is provided to \verb|ace.fn|, the base can provide a default target (here, \verb|ace.python.python|); 2) a concrete function can be generated immediately by providing a type assignment after \verb|ace.fn| using braces. Listing \ref{datatypest} generates Listing \ref{datatypes}.

Lines \ref{datatypest}.3-\ref{datatypest}.11 define a function that generates a recursive algebraic datatype representing a tree given a name and another Ace type for the data at the leaves. This type implemented by the active type family \verb|fp.Datatype|. Case names and their types are provided programmatically on lines \ref{datatypest}.5-\ref{datatypest}.10. These lines also show two more active type families: units (the type containing just one value), and immutable pairs. Line \ref{datatypest}.13 calls this function with the Ace type for dynamic Python values, \verb|py.dyn|, to generate a type, aliased \verb|DT|. This type is implemented using class inheritance when the target is Python, as seen on lines \ref{datatypes}.1-\ref{datatypes}.13. For C-like targets, a \verb|union| type can be used (not shown).

The generic function \verb|depth_gt_2| demonstrates two features. First, the base has been setup to treat the final expression in a top-level branch as the return value of the function, consistent with typical functional languages. Second, the \verb|case| ``method'' on line \ref{datatypest}.17 creatively reuses the syntax for dictionary literals to express a nested pattern matching construct. The patterns to the left of the colons are not treated as expressions (that is, a type and translation is never recursively assigned to them). The active type implements the standard algorithm for ensuring that the cases cover all possibilities and are not redundant. If the final case were omitted, for example, the algorithm would statically indicate an error, just as in statically-typed functional languages like ML and Haskell.

@@

\subsection{Product Types}
Data structures like structs, unions, records and objects\todo{mention that these are product types} can all be implemented as Ace type families  parameterized by a dictionary mapping field names to types. Each family differs slightly in the semantics of the operations available. Structs support mutation, for example, while records do not. Object systems typically introduce additional logic for calling methods, binding special variables like \verb|this|, and accessing information through an inheritance hierarchy. Listing \ref{ooclxfppy} shows an example of each of these. The prototypic object system delegates to a backing record if a field is not available in the struct. This example also demonstrates cross-compilation to C99, supported for a subset of operations in \verb|clx|. All three abstractions are implemented as simple \verb|struct|s, as can be seen in Listing \ref{ooclxfpc}. This example also demonstrates the use of argument annotations to generate concrete functions (line \ref{ooclxfppy}.9), which is a feature only available when running Ace in Python 3.

\subsection{Distributed Programming}
Many distributed programming abstractions can be understood as implementing object models with complex forms of dynamic dispatch. For example, in Charm++, dynamic dispatch involves message passing over a dynamically load-balanced network \cite{kale1993charm++}. While Charm++ is a sophisticated system, and we do not anticipate that implementing it as an Ace extension would be easy, it is possible to do so by targeting a system like Adaptive MPI, which exposes the Charm++ runtime system \cite{kale2009charm++}. %Erlang and other message-passing and actor-based systems can also be considered in this manner -- processes can be thought of as objects, channels as methods and messages as calls.

@@

A number of recent languages designed for distributed programming on clusters use a partitioned global address space model, including UPC \cite{upc}, Chapel \cite{chapel} and others. These languages provide first-class support for accessing data transparently across a massively parallel cluster. Their type systems track information about \emph{data locality} so that the compiler can emit more efficient code. The extension mechanism can also track this information in a manner analogous to how address space information is tracked in the OpenCL example above, and target an existing portable run-time such as GASNet \cite{bonachea2002gasnet}.

@@

%Type-Specific Literals && Wyvern \cite{globaldsl13} && \Circle && \CIRCLE && \Circle && \CIRCLE && \Circle \\
\end{longtable}
\caption{Comparison to related approaches to language-internal extensibility.}
\twocolumn
\end{figure*}
%\subsection{Active Libraries}

Libraries that contain compile-time logic have been called {\it active libraries} in prior proposals \cite{activelibraries}. A number of  projects, such as Blitz++, have taken advantage of the C++ preprocessor and template-based metaprogramming system to implement domain-specific optimizations \cite{veldhuizen2000blitz++}. In Ace, we replace these brittle mini-languages with a general-purpose language, Python, and significantly expand the notion of active libraries by consideration of types, base semantics and target languages as values (in this case, objects) in the compile-time metalanguage.

Generic functions are a novel strategy for {\it function polymorphism} -- defining functions that operate over more than a single type. In Ace, generic functions are implicitly polymorphic and can be called with arguments of {\it any type that supports the operations used by the function}. This approach is related to structural polymorphism \cite{malayeri2009structural}. Structural types make explicit the requirements on a function, unlike generic functions, which are only given singleton types because the requirements can be defined arbitrarily by extensions.
Structural typing can be compared to the more \emph{ad hoc} approach taken by dynamically-typed languages like Python itself, sometimes called ``duck typing''. It is also comparable to the C++ template system, as discussed previously. Debugging type errors can be more difficult when there are many nested generic functions that were implicitly given type assignments. Type debuggers like \verb|scalad| could help make this easier \cite{scalad}.

{\it Operator overloading} \cite{vanWijngaarden:Mailloux:Peck:Koster:Sintzoff:Lindsey:Meertens:Fisker:acta:1975} and {\it metaobject dispatch} \cite{Kiczales91} are run-time protocols that translate operator invocations into function calls. The function is typically selected according to the type or value of one or more operands. These protocols share the notion of {\it inversion of control} with our strategy. However, our strategy is a {\it compile-time} protocol where the typechecking and translation semantics are implemented for different operators. Our dispatch protocol was designed to be conceptually similar to Python's operator overloading protocol \cite{python}.

There are several other language-internal extension mechanisms that have been described in the literature. These differ from our mechanism based primarily on active types in one of the following ways, summarized in the table in Figure 1:
\begin{itemize}
\item Our mechanism is itself implemented as a library, rather than as a language extension.
\item Our mechanism reuses an existing language's syntax, rather than offering syntax extension support. As we saw in, for example, Listing \ref{datatypest}, this can still be reasonably natural. However, some mechanisms provide true syntax extensions.
\item Our mechanism permits true type system extensions. New types are not merely aliases, nor must their rules be admissible in some base type system. Term rewriting and macro systems do not handle types at all.
\item As described in Sec. \ref{safety}, our mechanism permits compositional reasoning. Techniques that allow global extensions to the syntax or semantics (e.g. SugarJ or rule injection systems like Xroma or Typed Racket) allow extensions so much control that they can interfere with one another or make reasoning about code difficult.
\item Our mechanism allows users to control the target language of compilation from within libraries. Many other mechanisms are based fundamentally on rewriting mechanisms.
\end{itemize}

When the mechanisms available in an existing language prove insufficient, researchers and domain experts often design a new language. A number of tools have been developed to assist with this task, including compiler generators, language workbenches and domain-specific language frameworks (cf \cite{fowler2010domain}). Extensible compilers can be considered a form of language framework as well due to the interoperability issues that relying on a compiler-specific extensions can introduce. It is difficult or impossible for these language-external approaches to achieve interoperability, because languages are not aware of each other's semantics.

@@

Ace has some limitations at the moment. Debuggers and other tools that rely not just on Python's syntax but also its semantics cannot be used directly, so if code generation introduces significant complexity that leads to bugs, this can be an issue. The OpenCL library we have implemented is a reasonably straightforward internalization of OpenCL itself, however, so debugging has not been a problem thusfar. We believe that active types can be used to control debugging, and plan to explore this in the future. We have also not yet evaluated the feasibility of implementing more advanced type systems (e.g. linear type systems, dependent type systems, flow-dependent type systems) using Ace.

Not all extensions will be useful. Indeed, some language designers worry that offering too much flexibility to users leads to abuse (this is, for example, widely credited as the reason why Java doesn't support operator overloading). We do not argue with this point. Instead, we argue that the potential for abuse must be balanced with the possibilities made available by a vibrant ecosystem of competing statically-typed abstractions that can be developed and deployed as libraries, and thus more easily evaluated in the wild. With an appropriate community process, this could lead to convergence via collections of curated, high-quality collections more quickly than is possible today, where most widely-adopted languages evolve quite slowly. %We also anticipate that coding guidelines and tools mandating the use of abstractions that are known to have certain desirable properties will replace designer-mandated enforcement.

This paper is presented as a language design paper. The theoretical foundations of this work lie in type-level computation. We have developed a simplified, type-theoretic formalism of active typechecking and translation, where we prove the safety properties we have outlined here, as well as several additional ones that are only possible to achieve using a typed metalanguage. That work is currently in submission at ESOP 2014. The work described here is original and extends that mechanism in several directions: Ace supports a rich syntax, makes syntax trees available to extensions, supports a pluggable backend and per-function base semantics. It is implemented in an existing language as a library and supports a wide range of eminently practical extensions. We hope that it will serve as a useful tool both to researchers and in practice. The mechanisms described here can be supported be any language that offers some form of quotation of function ASTs and a reasonably flexible collection of syntactic forms. They are particularly useful when the language has a variety of interoperability layers with other languages, like Python does.

>>>>

%\category{D.3.2}{Programming Languages}{Language Classifications}[Extensible Languages]
%\category{D.3.4}{Programming Languages}{Processors}[Compilers]
%\category{F.3.1}{Logics \& Meanings of Programs}{Specifying and Verifying and Reasoning about Programs}[Specification Techniques]
%\keywords
%type-level computation, typed compilation
\section{Introduction}\label{intro}
Asking programmers to import a new library is far more practical than asking them to  adopt a new programming language. Indeed, recent empirical studies underscore  the difficulties of driving new languages into adoption, finding that extrinsic factors like compatibility with existing codebases and libraries, team familiarity and tool support are at least as important as intrinsic factors \cite{Meyerovich:2013:EAP:2509136.2509515,chen05,nguyen2010survey}. As a result, many developers cannot use  abstractions they might prefer because these abstractions are only available in languages they cannot adopt \cite{Meyerovich:2012:SDR:2414721.2414724,Meyerovich:2013:EAP:2509136.2509515}. This issue was perhaps most succinctly expressed by a participant in a recent study by Basili et al. \cite{basili2008understanding} who stated ``I hate MPI, I hate C++. [But] if I had to choose again, I would probably choose the same.''

@@

\subsection{Recursive Labeled Sums}
\begin{codelisting}
\lstinputlisting{datatypes_t.py}
\caption{\texttt{[datatypes\_t.py]} An example using statically-typed functional datatypes.}
\label{datatypest}
\end{codelisting}
\begin{codelisting}
\lstinputlisting{datatypes.py}
\caption{\texttt{[datatypes.py]} The dynamically-typed Python code generated by running \texttt{acec datatypes\_t.py}.}
\label{datatypes}
\end{codelisting}
Listing \ref{datatypest} shows an example of the use of statically-typed functional datatypes (a.k.a. recursive labeled sum types) together with the Python implementation just described. It shows two syntactic conveniences that were not mentioned previously: 1) if no target is provided to \verb|ace.fn|, the base can provide a default target (here, \verb|ace.python.python|); 2) a concrete function can be generated immediately by providing a type assignment after \verb|ace.fn| using braces. Listing \ref{datatypest} generates Listing \ref{datatypes}.

Lines \ref{datatypest}.3-\ref{datatypest}.11 define a function that generates a recursive algebraic datatype representing a tree given a name and another Ace type for the data at the leaves. This type implemented by the active type family \verb|fp.Datatype|. A name for the datatype and the case names and types are provided programmatically on lines \ref{datatypest}.3-\ref{datatypest}.8. To support recursive datatypes, the case names are enclosed within a lambda term that will be passed a reference to the datatype itself. These lines also show two more active type families: units (the type containing just one value), and immutable pairs. Line \ref{datatypest}.13 calls this function with the Ace type for dynamic Python values, \verb|py.dyn|, to generate a type, aliased \verb|DT|. This type is implemented using class inheritance when the target is Python, as seen on lines \ref{datatypes}.1-\ref{datatypes}.13. For C-like targets, a \verb|union| type can be used (not shown).

The generic function \verb|depth_gt_2| demonstrates two features. First, the base has been setup to treat the final expression in a top-level branch as the return value of the function, consistent with typical functional languages. Second, the \verb|case| ``method'' on line \ref{datatypest}.17 creatively reuses the syntax for dictionary literals to express a nested pattern matching construct. The patterns to the left of the colons are not treated as expressions (that is, a type and translation is never recursively assigned to them). Instead, the active type implements the standard algorithm for ensuring that the cases cover all possibilities and are not redundant \cite{pfpl}. If the final case were omitted, for example, the algorithm would statically indicate an error, just as in statically-typed functional languages like ML.

@@

\subsection{Product Types}
Product types (types that group heterogeneous values together) like structs, unions, tuples, records and objects can all be implemented as Ace type families  parameterized by a dictionary mapping field names (indices, in the case of tuples) to types. Each family differs slightly in the semantics of the operations available. Structs support mutation, for example, while records do not. Object systems typically introduce additional logic for calling methods, binding special variables like \verb|this|, and accessing information through an inheritance hierarchy. Listing \ref{ooclxfppy} shows an example of each of these. The prototypic object system delegates to a backing record if a field is not available in the foreground. This example also demonstrates cross-compilation to C99, supported for a subset of operations in \verb|clx|. All three abstractions are implemented as simple \verb|struct|s, as can be seen in Listing \ref{ooclxfpc}. This example further demonstrates the use of argument annotations to generate concrete functions (line \ref{ooclxfppy}.9), which is a feature only available when using Python 3.3+.

\subsection{Distributed Programming}
Many distributed programming abstractions can be understood as implementing object models with complex forms of dynamic dispatch. For example, in Charm++, dynamic dispatch involves message passing over a dynamically load-balanced network \cite{kale2009charm++}. While Charm++ is a sophisticated system, and we do not anticipate that implementing it as an Ace extension would be easy, it is possible to do so by targeting a system like Adaptive MPI, which exposes the Charm++ runtime system \cite{kale2009charm++}. %Erlang and other message-passing and actor-based systems can also be considered in this manner -- processes can be thought of as objects, channels as methods and messages as calls.

@@

A number of recent languages designed for distributed programming on clusters use a partitioned global address space model, e.g. Chapel \cite{chapel}. These languages provide first-class support for accessing data transparently across a massively parallel cluster. Their type systems track information about \emph{data locality} so that the compiler can emit more efficient code. The extension mechanism can also track this information in a manner analogous to how address space information is tracked in the OpenCL example above, and target an existing portable run-time such as GASNet \cite{bonachea2002gasnet}.

@@

%Type-Specific Literals && Wyvern \cite{globaldsl13} && \Circle && \CIRCLE && \Circle && \CIRCLE && \Circle \\
\end{longtable}
\caption{Comparison to related approaches to language-internal extensibility.}\label{relatedtable}
\twocolumn
\end{figure*}
%\subsection{Active Libraries}

Some previous work has considered embedding statically-typed languages within other languages, including dynamically-typed languages. For example, Terra embeds a low-level statically-typed language within Lua \cite{terra}. These have not been fundamentally extensible, and generally require modifying the host language externally, leading to the ``bootstrapping'' problem of Sec. \ref{intro}.

Libraries containing compile-time logic have previously been called {\it active libraries} \cite{activelibraries}. A number of  projects, such as Blitz++, have taken advantage of C++ template metaprogramming to implement domain-specific optimizations \cite{veldhuizen2000blitz++}. Others have chosen custom metalanguages (e.g. Xroma \cite{activelibraries}). In Ace, we replace these brittle mini-languages with a general-purpose language, Python, and significantly expand the notion of active libraries by consideration of types, base semantics and target languages as values (in this case, objects) in the compile-time metalanguage.

Generic functions are a novel strategy for {\it function polymorphism} -- defining functions that operate over more than a single type. In Ace, generic functions are implicitly polymorphic and can be called with arguments of {\it any type that supports the operations used by the function}. This is related to structural polymorphism \cite{malayeri2009structural}. Structural types make explicit the requirements on a function, unlike generic functions, which are only given singleton types because the requirements can be defined arbitrarily by extensions.
Structural typing can be compared to the more \emph{ad hoc} approach taken by dynamically-typed languages like Python itself, sometimes called ``duck typing''. It is also comparable to the C++ template system, as discussed previously. Debugging type errors can be more difficult when there are many nested generic functions that were implicitly given type assignments. Type debuggers like \verb|scalad| could help make this easier \cite{scalad}.

{\it Operator overloading} \cite{vanWijngaarden:Mailloux:Peck:Koster:Sintzoff:Lindsey:Meertens:Fisker:acta:1975} and {\it metaobject dispatch} \cite{Kiczales91} are run-time protocols that translate operator invocations into function calls. The function is typically selected according to the type or value of one or more operands. These protocols share the notion of {\it inversion of control} with our strategy. However, our strategy is a {\it compile-time} protocol where the typechecking and translation semantics are implemented for different operators. %Our dispatch protocol was designed to be conceptually similar to Python's operator overloading protocol \cite{python}.

There are several other language-internal extension mechanisms that have been described in the literature. These differ from our mechanism in one of the following ways, summarized in Fig. \ref{relatedtable}:
\begin{itemize}
\item Our mechanism is itself implemented as a library, rather than as a language extension.
\item Our mechanism reuses an existing language's syntax, rather than offering syntax extension support. As we saw in, for example, Listing \ref{datatypest}, this can still be reasonably natural, and allows the language to benefit from a variety of existing tools.
\item Our mechanism permits true type system extensions. New types are not merely aliases, nor must their rules be admissible in some base type system.% Term rewriting and macro systems do not handle types at all.
\item As described in Sec. \ref{safety}, our mechanism permits compositional reasoning. Techniques that allow global extensions to the syntax or semantics (e.g. SugarJ or rule injection systems like Xroma or Typed Racket) allow extensions so much control that they can interfere with one another or make reasoning  difficult.
\item Our mechanism allows users to control the target language of compilation from within libraries. Many other mechanisms are based fundamentally on term-to-term rewriting.
\end{itemize}

When the mechanisms available in an existing language prove insufficient, researchers and domain experts often design a new language. A number of tools have been developed to assist with this task, including compiler generators, language workbenches and domain-specific language frameworks (cf \cite{erdweg2013state}). Extensible compilers can be considered a form of language framework as well due to the interoperability issues that relying on a compiler-specific extensions can introduce. It is difficult or impossible for these language-external approaches to achieve interoperability, because languages are not aware of each other's semantics and merging languages is not guaranteed to be sound.

@@

Ace has some limitations at the moment. Debuggers and other tools that rely not just on Python's syntax but also its semantics cannot be used directly, so if code generation introduces significant complexity that leads to bugs, this can be an issue. We believe that active types can be used to control debugging, and plan to explore this in the future. We have also not yet evaluated the feasibility of implementing more advanced type systems (e.g. linear, dependent or flow-dependent type systems) using Ace.

Not all extensions will be useful. Indeed, some language designers worry that offering too much flexibility to users leads to abuse (this is, for example, widely credited as the reason why Java doesn't support operator overloading). We do not argue with this point. Instead, we argue that the potential for abuse must be balanced with the possibilities made available by a vibrant ecosystem of competing statically-typed abstractions that can be developed and deployed as libraries, and thus more easily evaluated in the wild. With an appropriate community process, this could lead to a convergence toward stable collections of curated, high-quality collections more quickly than is possible today. %We also anticipate that coding guidelines and tools mandating the use of abstractions that are known to have certain desirable properties will replace designer-mandated enforcement.

This paper is presented as a language design paper. The theoretical foundations of this work lie in type-level computation. We have developed a simplified, type-theoretic formalism of active typechecking and translation, where we prove the safety properties we have outlined here, as well as several additional ones that are only possible to achieve using a typed metalanguage (currently in submission at ESOP 2014). The work described here  extends that mechanism in several directions: Ace supports a rich syntax, makes syntax trees available to extensions, supports a pluggable backend and per-function base semantics. It is implemented in an existing language and supports a wider range of practical extensions.

The mechanisms described here can be implemented within any language that offers some form of quotation of function ASTs and a reasonably flexible collection of syntactic forms and basic reflection facilities. It is particularly useful when the language supports interoperability layers with a number of target languages.
